---
title: "Tutorial: Using the Diarization Simulation Package (R + CLI)"
author: ""
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Tutorial: Using the Diarization Simulation Package (R + CLI)

This tutorial provides a documented example of use of the diarization-simulation package from R, calling the package's command-line interface via `system2()`. For background on the package, the statistical model, and installation instructions, see the README. For the equivalent Python workflow using the API directly, see tutorial.ipynb.

The worked example follows Gautheron et al. (2025) and addresses a concrete question: could an observed correlation between child vocalizations (CHI) and female adult vocalizations (FEM) be an artifact of diarization errors alone? The example has two parts — assessing sensitivity of R(CHI, FEM) to confusion errors, and testing whether an observed R is consistent with the null hypothesis that the true correlation is zero.

**Prerequisites**: The diarization-simulation package must be installed and on your PATH (see README). R packages used: **MASS** (multivariate normal) and base R. Code chunks that call the CLI are set to `eval = FALSE` so the document knits even if the simulator is not on your PATH; run those chunks interactively in RStudio (or set `eval = TRUE`) once the CLI is available.


## How to use this document

This tutorial is written as an R Markdown file, intended to be opened and run in RStudio, because we believe most of our readers will be more familiar with it than with a terminal. If you are not familiar with RStudio or R Markdown, Software Carpentry has beginner-friendly introductions to [RStudio](https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html), including to [R Markdown](https://swcarpentry.github.io/r-novice-gapminder/14-knitr-markdown.html). In a nutshell, R Markdown is a format that integrates written text and executable code; code sections are delimited by ```{r chunk-name} and ```. As you read through this tutorial, run each code chunk as you go by clicking the green play button in the top-right corner of the chunk — the output will appear either directly below the chunk or in a different pane in Rstudio (the console or "Plots" section).
Some chunks call the diarization-simulation package through the command line via `system2()`, because diarization-simulation is a Python package, not an R package. If you would prefer to use the command line directly instead, Software Carpentry has an introduction [here](https://swcarpentry.github.io/shell-novice/). Regardless of your approach, make sure the package is installed before running any chunks — see the README for instructions.

---

## Contents

1. [Quick start (5 minutes)](#quick-start-5-minutes)
2. [Why use this package?](#why-use-this-package)
3. [Installation and prerequisites](#installation-and-prerequisites)
4. [Complete worked example: R(CHI, FEM) — sensitivity and significance](#complete-worked-example-rchi-fem--sensitivity-and-significance)
5. [Step-by-step reference](#step-by-step-reference)
6. [References](#references)

---

## Quick start (5 minutes)

This section walks through the core workflow in three steps: creating a small example dataset, running the simulator on it, and inspecting the results. 

The chunk below creates a ground truth table with three recordings and saves it as a CSV file — you won't see any output yet, but after running it you should find a file called truth.csv in your working directory.

```{r quick-start, eval=FALSE}
# 1. Create a small ground truth CSV
truth <- data.frame(
  observation = 1:3,
  CHI  = c(120, 90, 150),
  OCH  = c(30, 15, 25),
  FEM  = c(200, 180, 220),
  MAL  = c(50, 70, 45)
)
write.csv(truth, "truth.csv", row.names = FALSE)

```

This next chunk passes the created truth.csv file to the simulator, which will run and print some progress information in the RStudio console. It may take a few seconds.

```{r run-sims}
# 2. Run the simulator (CLI); adjust path to diarization-simulate if needed
system2("diarization-simulate", c(
  "--truth", "truth.csv",
  "--output", "simulated.csv",
  "--algo", "vtc",
  "--samples", "100",
  "--distribution", "poisson",
  "--seed", "42"
))

```

Finally, the next chunk loads the simulator's output and prints the size of the output table, the first few rows, and the mean simulated counts per observation. The dimensions, in this example, will be 300: Each recording appears once per sample, so with 100 samples you should see 300 rows in total. Next, you can inspect the printed rows to get an idea of what the output file columns look like, but also the fact that the same 3 observations are repeated across samples, each time with slightly different simulated counts. This is what the simulator does, reflect the variability in observed counts due to algorithm variability and confusion.

```{r}
# 3. Load and inspect in R
sim <- read.csv("simulated.csv")
dim(sim)


head(sim, 20)

# Mean simulated counts per observation
aggregate(cbind(CHI, OCH, FEM, MAL) ~ observation, data = sim, FUN = mean)
```



## Complete worked example: R(CHI, FEM) — sensitivity and significance

This section illustrates two distinct ways to use the package, both centered on the same quantity: the correlation between child and female-adult vocalization counts, R(CHI, FEM). Although the two use cases use our simulation package, they answer different questions and would typically arise at different stages of an analysis.

The first use case is **sensitivity analysis**: given a _known true correlation_, how much does the algorithm distort the measured R(CHI, FEM)? This is useful when you want to understand how much your results could be affected by classification errors prior to looking at your own data. To do this, we construct a synthetic ground truth dataset where the true correlation between CHI and FEM is fixed at a known level and free of any noise — this is the "real truth". We then simulate what the algorithm would measure, and observe how its errors warp the vocalization counts, shifting the measured correlation away from the true one.


The second use case is **significance testing**: given an _observed R(CHI, FEM) from your own data_, could it have arisen from classification errors alone, even if the true underlying correlation is zero? This is useful when you have already obtained a result and want to assess whether it reflects a genuine relationship or is a plausible artifact of the algorithm. To do this, we construct a synthetic ground truth dataset where the true correlation is fixed at zero — again, free of any noise — and simulate what the algorithm would measure across many samples. This gives us a null distribution of measured correlations that we would expect to see if no true relationship existed. We then check whether the observed R(CHI, FEM) falls within or outside that distribution.



### Step 1: Examine the sensitivity of R(CHI, FEM) to confusion errors

We generate ground truth with a **known true correlation** (0 and ~0.3), simulate VTC via the CLI, then compute the **measured** R(CHI, FEM) in R and compare to the true R.

#### 1.1 Generate ground truth from a multivariate log-normal (in R)

We draw vocalization counts from a **multivariate log-normal** plus Poisson: log(counts) ~ MVN(μ, Σ), then counts = Poisson(exp(Z)). Parameters match the Python tutorial so results are comparable.

```{r make-ground-truth}
library(MASS)

SPEAKERS <- c("CHI", "OCH", "FEM", "MAL")

make_ground_truth <- function(n_obs, mean_log, std_log, corr_matrix, seed = 42) {
  set.seed(seed)
  D <- diag(std_log)
  cov_log <- D %*% corr_matrix %*% D
  Z <- MASS::mvrnorm(n_obs, mean_log, cov_log)
  rates <- exp(Z)
  counts <- matrix(rpois(length(rates), c(rates)), nrow = n_obs, ncol = 4)
  counts <- pmax(counts, 0L)
  df <- data.frame(observation = seq_len(n_obs), counts)
  colnames(df) <- c("observation", SPEAKERS)
  df
}

# Typical counts (geometric means): CHI ~1000, OCH ~300, FEM ~2000, MAL ~500
mean_log <- log(c(1000, 300, 2000, 500))
std_log <- c(0.4, 0.5, 0.35, 0.5)

# True R = 0: CHI and FEM uncorrelated (order: CHI, OCH, FEM, MAL; indices 1 and 3)
corr_r0 <- diag(4)
corr_r0[1, 3] <- corr_r0[3, 1] <- 0

# True R ≈ 0.3: positive CHI–FEM correlation on log scale
corr_r03 <- diag(4)
corr_r03[1, 3] <- corr_r03[3, 1] <- 0.35

n_obs <- 64
truth_r0   <- make_ground_truth(n_obs, mean_log, std_log, corr_r0)
truth_r03  <- make_ground_truth(n_obs, mean_log, std_log, corr_r03)

# True R(CHI, FEM)
cat("True R(CHI,FEM) when CHI–FEM uncorrelated:", cor(truth_r0$CHI, truth_r0$FEM), "\n")
cat("True R(CHI,FEM) when CHI–FEM correlated:  ", cor(truth_r03$CHI, truth_r03$FEM), "\n")

write.csv(truth_r0,  "example_truth_r0.csv",  row.names = FALSE)
write.csv(truth_r03, "example_truth_r03.csv", row.names = FALSE)
```

Helper to compute R(CHI, FEM) per sample (used below):

```{r measured-r-helper}
measured_r_per_sample <- function(sim_df) {
  samples <- unique(sim_df$sample)
  vapply(samples, function(s) {
    sub <- sim_df[sim_df$sample == s, ]
    cor(sub$CHI, sub$FEM)
  }, numeric(1))
}
```

#### 1.2 Run the simulator via CLI and compute measured R per sample

Call `diarization-simulate` twice (once per ground truth), then in R compute the Pearson correlation between CHI and FEM for each sample.

```{r simulate-and-r-per-sample, eval=FALSE}
# Simulate under true R = 0 (500 samples)
system2("diarization-simulate", c(
  "--truth", "example_truth_r0.csv",
  "--output", "sim_r0.csv",
  "--algo", "vtc",
  "--samples", "500",
  "--distribution", "poisson",
  "--seed", "42"
))

# Simulate under true R ≈ 0.3 (500 samples)
system2("diarization-simulate", c(
  "--truth", "example_truth_r03.csv",
  "--output", "sim_r03.csv",
  "--algo", "vtc",
  "--samples", "500",
  "--distribution", "poisson",
  "--seed", "43"
))

# Load and compute R(CHI, FEM) per sample (measured_r_per_sample defined above)
sim_r0  <- read.csv("sim_r0.csv")
sim_r03 <- read.csv("sim_r03.csv")
R_measured_r0  <- measured_r_per_sample(sim_r0)
R_measured_r03 <- measured_r_per_sample(sim_r03)
```

For the rest of the tutorial we assume these objects exist. If you did not run the CLI, you can load pre-generated CSVs or run the chunks above.

#### 1.3 Compare measured R to true R (sensitivity)

```{r sensitivity-print}
if (file.exists("sim_r0.csv") && file.exists("sim_r03.csv")) {
  sim_r0  <- read.csv("sim_r0.csv")
  sim_r03 <- read.csv("sim_r03.csv")
  R_measured_r0  <- measured_r_per_sample(sim_r0)
  R_measured_r03 <- measured_r_per_sample(sim_r03)
  cat("When TRUE R(CHI,FEM) = 0:\n")
  cat("  Mean measured R:", mean(R_measured_r0), "\n")
  cat("  Std measured R: ", sd(R_measured_r0), "\n")
  cat("  (Spurious correlation from confusion alone)\n\n")
  cat("When TRUE R(CHI,FEM) ≈ 0.3:\n")
  cat("  Mean measured R:", mean(R_measured_r03), "\n")
  cat("  Std measured R: ", sd(R_measured_r03), "\n")
  cat("  (Bias: measured R vs true 0.3)\n")
} else {
  cat("Run the 'simulate-and-r-per-sample' chunk (with eval=TRUE) after installing the CLI to see sensitivity output.\n")
}
```

**Interpretation:** Even when the true correlation is **zero**, the measured R is typically **positive** (spurious correlation). When the true R is 0.3, the measured R is often **inflated**. So R(CHI, FEM) is **sensitive** to confusion errors.

---

### Step 2: Test whether an observed R is consistent with classification errors alone

**Scenario:** You have 64 recordings, ran VTC, and observe **R_observed = 0.39**. Could this arise only from classification errors (true R = 0)? We simulate under the null and compute a p-value.

#### 2.1 Assume corpus size and observed R

```{r observed-r}
n_obs_corpus <- 64
R_observed <- 0.39
```

#### 2.2 Generate null ground truth (true R = 0) and run simulator

```{r null-truth-and-sim, eval=FALSE}
truth_null <- make_ground_truth(n_obs_corpus, mean_log, std_log, corr_r0, seed = 123)
write.csv(truth_null, "example_truth_null.csv", row.names = FALSE)

system2("diarization-simulate", c(
  "--truth", "example_truth_null.csv",
  "--output", "sim_null.csv",
  "--algo", "vtc",
  "--samples", "2000",
  "--distribution", "poisson",
  "--seed", "999"
))

sim_null <- read.csv("sim_null.csv")
R_null <- measured_r_per_sample(sim_null)
```

#### 2.3 Compare R_observed to the null distribution and compute p-value

```{r pvalue}
if (file.exists("sim_null.csv")) {
  sim_null <- read.csv("sim_null.csv")
  R_null <- measured_r_per_sample(sim_null)
  p_one_tailed <- mean(R_null >= R_observed)
  p_two_tailed <- mean(abs(R_null) >= abs(R_observed))
  cat("R_observed =", R_observed, "\n")
  cat("Null distribution: mean =", mean(R_null), ", std =", sd(R_null), "\n")
  cat("One-tailed p-value (H0: true R <= 0):", p_one_tailed, "\n")
  cat("Two-tailed p-value (H0: true R = 0):", p_two_tailed, "\n")
} else {
  cat("Run the 'null-truth-and-sim' chunk (with eval=TRUE) to generate sim_null.csv, then re-run this chunk.\n")
}
```

**Interpretation:** A **small** p-value suggests the observed R is not easily explained by classification errors alone. A **large** p-value means the observed R is consistent with the null (spurious correlation from algorithm errors).

#### 2.4 Optional: plot null distribution and R_observed

```{r null-plot}
if (file.exists("sim_null.csv")) {
  if (!exists("R_null")) {
    sim_null <- read.csv("sim_null.csv")
    R_null <- measured_r_per_sample(sim_null)
  }
  hist(R_null, breaks = 50, freq = FALSE, col = "gray", border = NA,
       main = "Null distribution of R under true R=0",
       xlab = "Measured R(CHI, FEM)", ylab = "Density")
  abline(v = R_observed, col = "red", lwd = 2)
  abline(v = mean(R_null), col = "black", lty = 2)
  legend("topright", legend = c(
    sprintf("R_observed = %s", R_observed),
    sprintf("Mean null R = %.3f", mean(R_null))
  ), col = c("red", "black"), lty = c(1, 2), lwd = c(2, 1))
} else {
  plot(0, type = "n", xlab = "Measured R(CHI, FEM)", ylab = "Density",
       main = "Null distribution (run null-truth-and-sim chunk to generate)")
}
```

You can save the plot with `pdf("example_null_R_chi_fem.pdf")` … `dev.off()` or use `ggplot2` if preferred.

---

### Summary of the two steps

| Step | Goal | What you do |
|------|------|-------------|
| **1. Sensitivity** | See how much R(CHI, FEM) is distorted by confusion | Generate ground truth with known true R (0 and ~0.3), run CLI simulator, compute measured R per sample in R; compare mean(measured R) to true R. |
| **2. Significance** | Test if an observed R could be due to errors alone | Take corpus size and R_observed; generate null truth (true R=0), run CLI, get null distribution of measured R in R; p-value = proportion of null R ≥ R_observed (or two-tailed). |

---

## Step-by-step reference

### Ground truth format

CSV with columns: `observation`, `CHI`, `OCH`, `FEM`, `MAL` (non-negative integer counts). Create in R, export from a spreadsheet, or use `truth-simulate` from a ChildProject corpus (see README).

### Running the simulation (CLI from R)

Use `system2("diarization-simulate", c("--truth", path_to_truth, "--output", path_to_output, "--algo", "vtc", "--samples", "1000", "--distribution", "poisson", "--seed", "42"))`. All arguments from the [main tutorial](tutorial.md) apply; run `diarization-simulate --help` for the full list.

### Using simulated data in R

1. `read.csv("output.csv")` to load the simulated data.
2. For each `sample`, run the same analysis you would on real algorithm output (e.g. `cor(CHI, FEM)` or a regression).
3. Aggregate across samples and compare to the true values (sensitivity) or to the null distribution (significance).

---

## References

- **Package:** [LAAC-LSCP/diarization-simulation](https://github.com/LAAC-LSCP/diarization-simulation)
- **Manuscript:** Gautheron, L., Kidd, E., Malko, A., Lavechin, M., & Cristia, A. (2025). *Classification errors distort findings in automated speech processing: examples and solutions from child-development research.* [DOI: 10.31234/osf.io/u925y](http://dx.doi.org/10.31234/osf.io/u925y)

For installation, ground-truth generation from a ChildProject corpus, and the statistical model, see the project **README.md** and the [main tutorial](tutorial.md).
