---
title: "Tutorial: Using the Diarization Simulation Package (R + CLI)"
author: ""
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Tutorial: Using the Diarization Simulation Package (R + CLI)

This tutorial provides a documented example of use of the diarization-simulation package from R, calling the package's command-line interface via `system2()`. For background on the package, the statistical model, and installation instructions, see the README. For the equivalent Python workflow using the API directly, see tutorial.ipynb.

The worked example follows Gautheron et al. (2025) and addresses a concrete question: could an observed correlation between child vocalizations (CHI) and female adult vocalizations (FEM) be an artifact of diarization errors alone? The example has two parts — assessing sensitivity of R(CHI, FEM) to confusion errors, and testing whether an observed R is consistent with the null hypothesis that the true correlation is zero.

**Prerequisites**: The diarization-simulation package must be installed and on your PATH (see README). R packages used: **MASS** (multivariate normal) and base R. Code chunks that call the CLI are set to `eval = FALSE` so the document knits even if the simulator is not on your PATH; run those chunks interactively in RStudio (or set `eval = TRUE`) once the CLI is available.


## How to use this document

This tutorial is written as an R Markdown file, intended to be opened and run in RStudio, because we believe most of our readers will be more familiar with it than with a terminal. If you are not familiar with RStudio or R Markdown, Software Carpentry has beginner-friendly introductions to [RStudio](https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html), including to [R Markdown](https://swcarpentry.github.io/r-novice-gapminder/14-knitr-markdown.html). In a nutshell, R Markdown is a format that integrates written text and executable code; code sections are delimited by ```{r chunk-name} and ```. As you read through this tutorial, run each code chunk as you go by clicking the green play button in the top-right corner of the chunk — the output will appear either directly below the chunk or in a different pane in Rstudio (the console or "Plots" section).
Some chunks call the diarization-simulation package through the command line via `system2()`, because diarization-simulation is a Python package, not an R package. If you would prefer to use the command line directly instead, Software Carpentry has an introduction [here](https://swcarpentry.github.io/shell-novice/). Regardless of your approach, make sure the package is installed before running any chunks — see the README for instructions.

---

## Contents

1. [Quick start (5 minutes)](#quick-start-5-minutes)
2. [Why use this package?](#why-use-this-package)
3. [Installation and prerequisites](#installation-and-prerequisites)
4. [Complete worked example: R(CHI, FEM) — sensitivity and significance](#complete-worked-example-rchi-fem--sensitivity-and-significance)
5. [Step-by-step reference](#step-by-step-reference)
6. [References](#references)

---

## Quick start (5 minutes)

This section walks through the core workflow in three steps: creating a small example dataset, running the simulator on it, and inspecting the results. 

The chunk below creates a ground truth table with three recordings and saves it as a CSV file — you won't see any output yet, but after running it you should find a file called truth.csv in your working directory.

```{r quick-start}
# 1. Create a small ground truth CSV
truth <- data.frame(
  observation = 1:3,
  CHI  = c(120, 90, 150),
  OCH  = c(30, 15, 25),
  FEM  = c(200, 180, 220),
  MAL  = c(50, 70, 45)
)
write.csv(truth, "truth.csv", row.names = FALSE)

```

This next chunk passes the created truth.csv file to the simulator, which will run and print some progress information in the RStudio console. It may take a few seconds.

```{r run-sims}
# 2. Run the simulator (CLI); adjust path to diarization-simulate if needed
system2("diarization-simulate", c(
  "--truth", "truth.csv",
  "--output", "simulated.csv",
  "--algo", "vtc",
  "--samples", "100",
  "--distribution", "poisson",
  "--seed", "42"
))

```

Finally, the next chunk loads the simulator's output and prints the size of the output table, the first few rows, and the mean simulated counts per observation. The dimensions, in this example, will be 300: Each recording appears once per sample, so with 100 samples you should see 300 rows in total. Next, you can inspect the printed rows to get an idea of what the output file columns look like, but also the fact that the same 3 observations are repeated across samples, each time with slightly different simulated counts. This is what the simulator does, reflect the variability in observed counts due to algorithm variability and confusion.

```{r}
# 3. Load and inspect in R
sim <- read.csv("simulated.csv")
dim(sim)


head(sim, 20)

# Mean simulated counts per observation
aggregate(cbind(CHI, OCH, FEM, MAL) ~ observation, data = sim, FUN = mean)
```



## Complete worked example: R(CHI, FEM) — sensitivity and significance

This section illustrates two distinct ways to use the package, both centered on the same quantity: the correlation between child and female-adult vocalization counts, R(CHI, FEM). Although the two use cases use our simulation package, they answer different questions and would typically arise at different stages of an analysis.

The first use case is **sensitivity analysis**: given a _known true correlation_, how much does the algorithm distort the measured R(CHI, FEM)? This is useful when you want to understand how much your results could be affected by classification errors prior to looking at your own data. To do this, we construct a synthetic ground truth dataset where the true correlation between CHI and FEM is fixed at a known level and free of any noise — this is the "real truth". We then simulate what the algorithm would measure, and observe how its errors warp the vocalization counts, shifting the measured correlation away from the true one.


The second use case is **significance testing**: given an _observed R(CHI, FEM) from your own data_, could it have arisen from classification errors alone, even if the true underlying correlation is zero? This is useful when you have already obtained a result and want to assess whether it reflects a genuine relationship or is a plausible artifact of the algorithm. To do this, we construct a synthetic ground truth dataset where the true correlation is fixed at zero — again, free of any noise — and simulate what the algorithm would measure across many samples. This gives us a null distribution of measured correlations that we would expect to see if no true relationship existed. We then check whether the observed R(CHI, FEM) falls within or outside that distribution.



### Use case 1: Examine the sensitivity of R(CHI, FEM) to confusion errors

In this section we create two synthetic ground truth datasets — one where CHI and FEM are uncorrelated (true R = 0), and one where they have a moderate positive correlation (true R ≈ 0.3) — run the VTC simulator on each, and compare the measured R(CHI, FEM) to the known true value. This lets us see directly how much the algorithm inflates or distorts the correlation.


#### 1.1 Generate ground truth from a multivariate log-normal (in R)

The chunk below generates the two ground truth datasets and saves them as CSV files. Vocalization counts are drawn from a multivariate log-normal distribution: we first sample log-counts from a multivariate normal (which lets us control the correlation structure), then exponentiate and pass through a Poisson to get integer counts. This is a standard way to generate realistic, non-negative count data with a known correlation structure. The parameters are chosen to reflect typical vocalization counts in daylong recordings, and match those used in the Python tutorial so results are comparable across the two formats.

After running this chunk, you should see the true R(CHI, FEM) for each dataset printed in the console. Note that the correlation on the count scale will be lower than the value specified in the log-normal model (0.35), because the exponentiation and Poisson sampling both attenuate correlations — so expect to see something closer to 0.15 than 0.3 for the correlated dataset, and something close to 0 for the uncorrelated one. This is expected behaviour. You will also find two new CSV files in your working directory: example_truth_r0.csv and example_truth_r03.csv.

```{r make-ground-truth}
library(MASS)

SPEAKERS <- c("CHI", "OCH", "FEM", "MAL")

make_ground_truth <- function(n_obs, mean_log, std_log, corr_matrix, seed = 42) {
  set.seed(seed)
  D <- diag(std_log)
  cov_log <- D %*% corr_matrix %*% D
  Z <- MASS::mvrnorm(n_obs, mean_log, cov_log)
  rates <- exp(Z)
  counts <- matrix(rpois(length(rates), c(rates)), nrow = n_obs, ncol = 4)
  counts <- pmax(counts, 0L)
  df <- data.frame(observation = seq_len(n_obs), counts)
  colnames(df) <- c("observation", SPEAKERS)
  df
}

# Typical counts (geometric means): CHI ~1000, OCH ~300, FEM ~2000, MAL ~500
mean_log <- log(c(1000, 300, 2000, 500))
std_log <- c(0.4, 0.5, 0.35, 0.5)

# True R = 0: CHI and FEM uncorrelated (order: CHI, OCH, FEM, MAL; indices 1 and 3)
corr_r0 <- diag(4)
corr_r0[1, 3] <- corr_r0[3, 1] <- 0

# True R ≈ 0.3: positive CHI–FEM correlation on log scale
corr_r03 <- diag(4)
corr_r03[1, 3] <- corr_r03[3, 1] <- 0.35

n_obs <- 64
truth_r0   <- make_ground_truth(n_obs, mean_log, std_log, corr_r0)
truth_r03  <- make_ground_truth(n_obs, mean_log, std_log, corr_r03)

# True R(CHI, FEM)
cat("True R(CHI,FEM) when CHI–FEM uncorrelated:", cor(truth_r0$CHI, truth_r0$FEM), "\n")
cat("True R(CHI,FEM) when CHI–FEM correlated:  ", cor(truth_r03$CHI, truth_r03$FEM), "\n")

write.csv(truth_r0,  "example_truth_r0.csv",  row.names = FALSE)
write.csv(truth_r03, "example_truth_r03.csv", row.names = FALSE)
```



#### 1.2 Run the simulator and compute measured correlation per sample

This chunk does three things: it runs the simulator twice (once for each ground truth dataset), loads the results, and computes the Pearson correlation between CHI and FEM separately for each of the 500 simulated samples. This may take a moment to run — you should see some progress output in the console while the simulator is working.

Once it completes, you will have four new objects in your R environment: sim_r0 and sim_r03 (the raw simulated counts), and R_measured_r0 and R_measured_r03 (a vector of 500 correlations, one per sample, for each dataset).

```{r simulate-and-r-per-sample}
# Simulate under true R = 0 (500 samples)
system2("diarization-simulate", c(
  "--truth", "example_truth_r0.csv",
  "--output", "sim_r0.csv",
  "--algo", "vtc",
  "--samples", "500",
  "--distribution", "poisson",
  "--seed", "42"
))

# Simulate under true R ≈ 0.3 (500 samples)
system2("diarization-simulate", c(
  "--truth", "example_truth_r03.csv",
  "--output", "sim_r03.csv",
  "--algo", "vtc",
  "--samples", "500",
  "--distribution", "poisson",
  "--seed", "43"
))

# Load and compute R(CHI, FEM) per sample (measured_r_per_sample defined above)
sim_r0  <- read.csv("sim_r0.csv")
sim_r03 <- read.csv("sim_r03.csv")

# Helper function to compute R(CHI, FEM) per sample
measured_r_per_sample <- function(sim_df) {
  samples <- unique(sim_df$sample)
  vapply(samples, function(s) {
    sub <- sim_df[sim_df$sample == s, ]
    cor(sub$CHI, sub$FEM)
  }, numeric(1))
}

R_measured_r0  <- measured_r_per_sample(sim_r0)
R_measured_r03 <- measured_r_per_sample(sim_r03)
```


#### 1.3 Compare measured correlation to true correlation (sensitivity)

This chunk prints the mean and standard deviation of the measured R(CHI, FEM) across the 500 samples, for each dataset. You should expect to see that even when the true correlation is zero, the mean measured R is noticeably positive — this is the spurious correlation introduced by algorithm confusion errors. For the correlated dataset, the mean measured R will differ from the true count-scale correlation we observed in section 1.1, reflecting how the algorithm inflates it further.

The standard deviation across samples tells you how much variability there is in the measured R from one simulated run to the next — a useful indicator of how stable the bias is.

```{r sensitivity-print}
if (file.exists("sim_r0.csv") && file.exists("sim_r03.csv")) {
  sim_r0  <- read.csv("sim_r0.csv")
  sim_r03 <- read.csv("sim_r03.csv")
  R_measured_r0  <- measured_r_per_sample(sim_r0)
  R_measured_r03 <- measured_r_per_sample(sim_r03)
  cat("When TRUE R(CHI,FEM) = 0:\n")
  cat("  Mean measured R:", mean(R_measured_r0), "\n")
  cat("  Std measured R: ", sd(R_measured_r0), "\n")
  cat("  (Spurious correlation from confusion alone)\n\n")
  cat("When TRUE R(CHI,FEM) ≈ 0.3:\n")
  cat("  Mean measured R:", mean(R_measured_r03), "\n")
  cat("  Std measured R: ", sd(R_measured_r03), "\n")
  cat("  (Measured R given true count-scale R ≈ 0.15, and underlying R ≈ .3)\n")
} else {
  cat("Run the 'simulate-and-r-per-sample' chunk first!\n")
}
```

**Interpretation:** Even when the true correlation is zero, the measured R is on average around 0.31 — a substantial spurious correlation introduced by confusion errors alone. Strikingly, the measured R is almost identical when the true correlation is ~0.15, suggesting that the algorithm's errors dominate the signal to the point where the measured R tells us very little about the true underlying relationship. This is a strong illustration of why R(CHI, FEM) is sensitive to confusion errors: the measured value is heavily shaped by the algorithm's behaviour regardless of the ground truth. 

---

### Use case 2: Test whether an observed R is consistent with classification errors alone

**Scenario:** You have 64 recordings, ran VTC, and observe R(CHI, FEM) = 0.39. Given what we learned in Use case 1 — that confusion errors alone produce a mean measured R of around 0.31 — this observed value may not be as striking as it first appears. The question is whether it falls within the range of what we would expect to see under the null hypothesis that the true correlation is zero, or whether it is extreme enough to suggest a genuine relationship.

#### 2.1 Set corpus size and observed R

This chunk simply stores the two values we will use throughout this section. You can also adjust R_observed to match your own data if you are applying this to a real analysis.

```{r observed-r}
n_obs_corpus <- 64
R_observed <- 0.39
```

#### 2.2 Generate null ground truth (true R = 0) and run simulator

This chunk does two things in sequence: it first generates a new synthetic ground truth dataset with true R = 0 (using the same make_ground_truth() function from section 1.1, but with a different random seed), saves it as example_truth_null.csv, and then immediately passes it to the simulator. We use 2000 samples here — more than in Use case 1 — to get a stable null distribution. This will take longer to run than the previous simulation chunks. When it completes, R_null will be a vector of 2000 measured correlations representing what the algorithm would produce by chance alone.

```{r null-truth-and-sim}
truth_null <- make_ground_truth(n_obs_corpus, mean_log, std_log, corr_r0, seed = 123)
write.csv(truth_null, "example_truth_null.csv", row.names = FALSE)

system2("diarization-simulate", c(
  "--truth", "example_truth_null.csv",
  "--output", "sim_null.csv",
  "--algo", "vtc",
  "--samples", "2000",
  "--distribution", "poisson",
  "--seed", "999"
))

sim_null <- read.csv("sim_null.csv")
R_null <- measured_r_per_sample(sim_null)
```

#### 2.3 Compare R_observed to the null distribution and compute p-value

This chunk computes how often the null distribution produces a measured R as extreme as the observed value. The p-value is simply the proportion of null samples where the measured R meets or exceeds R_observed. Given that the null distribution is already centred around ~0.31, you should expect the p-value to be relatively large — meaning R_observed = 0.39 may well be consistent with confusion errors alone. A small p-value would suggest the observed R is genuinely difficult to explain by errors alone.

```{r pvalue}
if (file.exists("sim_null.csv")) {
  sim_null <- read.csv("sim_null.csv")
  R_null <- measured_r_per_sample(sim_null)
  p_one_tailed <- mean(R_null >= R_observed)
  p_two_tailed <- mean(abs(R_null) >= abs(R_observed))
  cat("R_observed =", R_observed, "\n")
  cat("Null distribution: mean =", mean(R_null), ", std =", sd(R_null), "\n")
  cat("One-tailed p-value (H0: true R <= 0):", p_one_tailed, "\n")
  cat("Two-tailed p-value (H0: true R = 0):", p_two_tailed, "\n")
} else {
  cat("Run the 'null-truth-and-sim' chunk (with eval=TRUE) to generate sim_null.csv, then re-run this chunk.\n")
}
```

**Interpretation:** A **small** p-value (below your chosen significance threshold, e.g. .05)) suggests the observed R is not easily explained by classification errors alone. 

#### 2.4 Plot null distribution and R_observed

```{r null-plot}
if (file.exists("sim_null.csv")) {
  if (!exists("R_null")) {
    sim_null <- read.csv("sim_null.csv")
    R_null <- measured_r_per_sample(sim_null)
  }
  hist(R_null, breaks = 50, freq = FALSE, col = "gray", border = NA,
       main = "Null distribution of R under true R=0",
       xlab = "Measured R(CHI, FEM)", ylab = "Density")
  abline(v = R_observed, col = "red", lwd = 2)
  abline(v = mean(R_null), col = "black", lty = 2)
  legend("topright", legend = c(
    sprintf("R_observed = %s", R_observed),
    sprintf("Mean null R = %.3f", mean(R_null))
  ), col = c("red", "black"), lty = c(1, 2), lwd = c(2, 1))
} else {
  plot(0, type = "n", xlab = "Measured R(CHI, FEM)", ylab = "Density",
       main = "Null distribution (run null-truth-and-sim chunk to generate)")
}
```

You can save the plot with `pdf("example_null_R_chi_fem.pdf")` … `dev.off()` or use `ggplot2` if preferred.




## References

- **Package:** [LAAC-LSCP/diarization-simulation](https://github.com/LAAC-LSCP/diarization-simulation)
- **Manuscript:** Gautheron, L., Kidd, E., Malko, A., Lavechin, M., & Cristia, A. (2025). *Classification errors distort findings in automated speech processing: examples and solutions from child-development research.* [DOI: 10.31234/osf.io/u925y](http://dx.doi.org/10.31234/osf.io/u925y)

For installation, ground-truth generation from a ChildProject corpus, and the statistical model, see the project **README.md**.
